Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-12-22 code/Rmd/ALL_REPROCESSED.Rmd
temp commit

---
title: "Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

>>>>

---
title: "Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and PCA
based model selection)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

### 2017-12-22 code/Rmd/ALL_REPROCESSED_v2.Rmd
temp commit

knit2synapse::knitToFolderEntity(file = "./ALL_REPROCESSED.Rmd",
                                 parentId = "syn8672415",
                                 entityName = 'ALL Reprocessed RNASeq (Normalised)')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn8672415';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of reprocessed rnaseq from all 7 brain regions';

thisFileName <- 'ALL_REPROCESSED.Rmd'

@@

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

@@

  # Estimate voom weights
  voomWeights = limma::voom(cnt[rownames(lcpm),colnames(lcpm)], design = designMat, plot=F)$weights

@@

                                # Perform weighted mixed linear models for every gene
                                expr = data.frame(Gx = lcpm[i,])
                                data = cbind(expr,
                                             designMat[rownames(expr),],
                                             covars[rownames(designMat),groupingVar,drop=F])
                                formula = paste(setdiff(colnames(data), c('Gx', groupingVar)), collapse = '+') %>%
                                  paste('Gx ~',.) %>%
                                  c(.,paste0('(1|',groupingVar,')')) %>%
                                  paste(collapse = '+')

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(Gene.ID = rownames(lcpm)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }

@@

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           c('individualIdentifier'), 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

@@

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -Gene.ID) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = (better/(better+worse)) * 100)

@@

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > 1) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(Gene.ID, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('BrainRegion.Diagnosis')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

### Normalisation
1. BrainRegion.Diagnosis is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect
```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)

# Estimate consensus correlation coefficient
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
                            design=DM1$design, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

@@

### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
Differential expression is performed on the BrainRegion.Diagnosis x Gender variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=7, fig.width=24}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$BrainRegion.Diagnosis.Gender = factor(paste(COVARIATES$BrainRegion.Diagnosis,
                                                       COVARIATES$Gender,
                                                       sep = '_'))

# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(adjust.covars, c('BrainRegion.Diagnosis', 'Gender'))), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

@@

# Re-estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

>>>>

knit2synapse::knitToFolderEntity(file = "./ALL_REPROCESSED_v2.Rmd",
                                 parentId = "syn8672415",
                                 entityName = 'ALL Reprocessed RNASeq (Normalised - sparse model)')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn8672415';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of reprocessed RNASeq from all 7 brain regions';

thisFileName <- 'ALL_REPROCESSED_v2.Rmd'

@@

### Normalisation (iterative design)
1. BrainRegion.Diagnosis is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

@@

  # Estimate voom weights
  cnt[is.na(cnt)] = 0
  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights

@@

                                # Perform weighted mixed linear models for every gene
                                expr = data.frame(Gx = lcpm[i,])
                                data = cbind(expr,
                                             designMat[rownames(expr),],
                                             Individual_ID = groupingVar)
                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }

@@

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           covars[colnames(cnt),'individualIdentifier'], 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

@@

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = better)

@@

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('BrainRegion.Diagnosis')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'individualIdentifier'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'individualIdentifier'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'individualIdentifier'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

@@

### Differential expression analysis (with BrainRegion.ApoE as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.ApoE', setdiff(postAdjustCovars, 'BrainRegion.Diagnosis')), drop = F],
                         Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

@@

# Re-estimate voom weights
VOOM.WEIGHTS = voom(NEW.COUNTS,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

# Fit contrast
cntr = c()
for (region in c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  tmp = combn(paste0('BrainRegion.ApoE', region, '_', c('0', '1', '2')),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.ApoE','',Comparison),
                Study = 'ALL',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

all.diff.exp = c(all.diff.exp, list(ApoE = DE))
all.fit = c(all.fit, list(ApoE = FIT))
```

### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
Differential expression is performed on the BrainRegion.Diagnosis x Gender variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp3, fig.height=7, fig.width=24}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$BrainRegion.Diagnosis.Gender = factor(paste(COVARIATES$BrainRegion.Diagnosis,
                                                       COVARIATES$Gender,
                                                       sep = '_'))

# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(postAdjustCovars, c('BrainRegion.Diagnosis', 'Gender'))), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(NEW.COUNTS,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

### 2017-12-22 code/Rmd/MAYO_REPROCESSED.Rmd
temp commit

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp3, fig.height=5, fig.width=18}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis', setdiff(postAdjustCovars, 'AgeAtDeath')), drop = F], 
                         Intercept = F)
DESIGN$design[,grep('BrainRegion.Diagnosis',colnames(DESIGN$design))] = DESIGN$design[,grep('BrainRegion.Diagnosis',colnames(DESIGN$design))] * COVARIATES[,'AgeAtDeath']
colnames(DESIGN$design)[grep('BrainRegion.Diagnosis',colnames(DESIGN$design))] = 
  paste(colnames(DESIGN$design)[grep('BrainRegion.Diagnosis',colnames(DESIGN$design))],'AOD', sep = '.')
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

>>>>

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp3, fig.height=5, fig.width=18, cache = FALSE}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis', setdiff(postAdjustCovars, 'AgeAtDeath')), drop = F], 
                         Intercept = F)
DESIGN$design[,grep('BrainRegion.Diagnosis',colnames(DESIGN$design))] = DESIGN$design[,grep('BrainRegion.Diagnosis',colnames(DESIGN$design))] * COVARIATES[,'AgeAtDeath']
colnames(DESIGN$design)[grep('BrainRegion.Diagnosis',colnames(DESIGN$design))] = 
  paste(colnames(DESIGN$design)[grep('BrainRegion.Diagnosis',colnames(DESIGN$design))],'AOD', sep = '.')
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
